Stage-Splitting a Modal Language

### 2014-09-28 paper/semantics.tex

\section {Staged Programming in \lang}
\label{sec:staging}

@@

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang clearly and unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each term with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  After splitting, those parts of the \lang\ term in stage \bbone\ will end up in the precomputation, 
and those parts in stage \bbtwo\ will end up in the residual.
Within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.  
Indeed this is a necessary property if we want to be able to split the term.

@@

As stated in \cref{sec:staging}, in this paper we describe splitting in the context of \lang,
a typed two-stage lambda calculus featuring products, sums, and isorecursive types.
The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
restricted to two stages and extended with general sums and recursion\footnote{This restriction to two stages is made primarily for 
simplicity of presentation.  All of these techniques could feasibly be extended to more
complicated stage systems.}.  The following subsections provide a brief overview of the type system for \lang, and highlight important differences in its semantics from this prior work.

@@

\subsection{Non-Duplicating First-Stage Evaluation}

@@

\subsection{Dynamics}
\label{ssec:dynamics}

@@

\subsection {A Partial Evaluation System}
\label{sec:partialeval}

>>>>

\section {Example: Staged Quickselect in \lang}
\label{sec:staging}

@@

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  After splitting, those parts of the \lang\ term in stage \bbone\ will end up in the precomputation, 
and those parts in stage \bbtwo\ will end up in the residual.
Typing rules for \lang ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.

@@

As stated in \cref{sec:staging}, we use \lang,
a typed two-stage lambda calculus featuring products, sums, and isorecursive types, as language to define and analyze the splitting transform.
The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
restricted to two stages and extended with general sums and recursion\footnote{This restriction to two stages is made primarily for 
simplicity of presentation.  All of these techniques could feasibly be extended to more
complicated stage systems.}.  The following subsections provide a brief overview of the type system for \lang, and highlight important differences in its semantics from this prior work.

@@

\subsubsection{Non-Duplicating First-Stage Evaluation}

@@

\subsubsection{Dynamics}
\label{ssec:dynamics}

@@

\subsubsection {A Partial Evaluation System}
\label{sec:partialeval}
